Regularising Lotka--Volterra Interaction Matrices by Species Grouping

### 2021-10-24 article.rnw
Correct spelling.

However, making any use of a model requires experimentally or observationally determining the values of each parameter in the model.
This poses a problem for species interacton models, due to the sheer number of parameters that must be determined.
Even the simplest models, with no environmental dependence, require one parameter for the interaction of each pair of species \parencite{novak2016characterizing}.
Hence the number of parameters scales quadratically with the number of species; over \num{100} parameters are required for a community of just \num{10} species.
Determining these parameters experimentally requires at least one experiment per parameter, and preferably more \parencite{hart2018quantify}.
This is expensive, and requires a lot of effort.
As such, these models are typically restricted to a small number of focal species \parencite{ovaskainen2017species}.
Real communities can contain hundreds of species \parencite{cornell1999unsaturation}, or even thousands for microbial communities assessed by genetic metabarcoding \parencite{warton2015variables}.
We need a way to reduce the number of parameters the models require, to regularise them.

@@

I propose that groups might be selected without regard to any \emph{a priori} definition, based purely upon which gives the most parsimonius model.
In this study, I investigate the usefulness and correctness of such a selection, using experimental \emph{Drosophila} fecundity data provided by \textcite{terry2021natural}.
I fit the Lotka--Volterra model, with grouped species, to find the grouping that gives the most parsimonious model.
I group species independently with regard to their effects upon other species and their responses to other species, to find whether the appropriate groupings might differ for these two cases.
Finally, I compare the resultant groupings with species phylogeny, to provide further investigation of the suitability of \emph{a priori} phylogenetic grouping.

@@

Species in the same column grouping were constrained to have the same competitive effect upon all other species, while species in the same column grouping were contrained to have the same competitive response to all other species.
Formally, if species $a$ and $b$ were in the same column grouping, $\alpha_{ia}=\alpha_{ib} \, \forall i$, and if $a$ and $b$ were in the same row grouping, $\alpha_{aj}=\alpha_{bj} \, \forall j$.
As such, the total number of free competition coefficients in the model was equal to the number of row groups times the number of column groups.
The species' instrinsic growth rates, $r_i$, were not affected by grouping; each species always had its own growth rate fitted independently of other species.

@@

To assess the parsimony of each moel, all \num{\Sexpr{numbers::bell(6)^2}} possible models were fitted.
As a measure of parsimony, I selected AIC over BIC on the basis that we want to find the best model for making ecological predictions, rather than seeking the ``true'' \parencite{aho2014model}.
Further, AIC will be more conservative in promoting grouping, as BIC penalises number of parameters more for any reasonably sized data set.
As such, I calculated the AIC for each model.

@@

\begin{figure}
	\centering
	\begin{forest} phylogeny
		[
			[
				[
					[\emph{D. simulans}]
					[\emph{D. birchii}]
				]
				[
					[\emph{D. pandora}]
					[\emph{D. pseudoananassae}]
				]
			]
			[
				[
					[\emph{D. pallidifrons}]
					[\emph{D. sulfurigaster}]
				]
			]
		]
	\end{forest}
	%TODO: Mark subgroups and subgenuses on the diagram.

@@

A high likelihood of coclassifcation was associated with phylogeny in one instance, but by no means consistently.
The phylogeny divides the six spcies into three most closely related pairs (\cref{fig:phylogeny}).
The pair of \emph{D. pallidifrons} and \emph{D. sulfurigaster}, which both belonged to a different subgenus than the other four species, were also the only pair of species to be coclassified with relatively high probability in both the row and column groupings.
In constrast, each of the other pairs were coclassified with very low probabilities in both the row and column groupings.
Furthermore, both row and column groupings gave high probabilities of coclassification to pairs of species split across the two subgenera in the study---the row groupings frequently coclassified \emph{D. sulfurigaster} and \emph{D. simulans}, while the column groupings frequently coclassified both \emph{D. pallidifrons} and \emph{D. sulfurigaster} of the Drosophila subgenus with \emph{D. birchii} of the Sophophora subgenus.

@@

This stands in contrast to the popular wisdom in ecology that species should form some sort of clear groups, be they the functional groups so often used \parencite{blondel2003guilds}, or taxonomic groups per a relationship between phylogeny and ecological niche \parencite{godoy2014phylogenetic}.
It also contrasts with previous studies that have succeeded in finding clear groups, such as \textcite{allesina2009food}, who used AIC to find the best grouping of species, but in a binary food web rather than a qualitative competitive context.
Perhaps species simply do not group to the same extent in a competitive context.
The results suggest that this is unlikely, as grouping was still likely to increase parsimony.
Further, there is no reason to suspect that speies would not have clear similarities or dissimilarities in their ecological niches, and therefore their competitive effects and responses.
Perhaps the lack of clarity in the groupings is instead a product of the study system chosen.

I chose a study system of \emph{Drosophila} partially on the basis that it would allow clear analysis of the species phylogenetic relationships.
Phylogeny proved a poor predictor of optimal grouping, as previous studies have suggested that it might \parencite{godoy2014phylogenetic,cahill2008does}.
A study system with species spanning multiple functional groups, however, might have shown clearer groupings.
While functional groups have previously shown to be poor predictors of competitive interactions \parencite{bret2008plant}, studies in other contexts have succeeded in isolating functional groups by selecting for model parsimony \parencite{allesina2009food}.
In further support of functional groups over phylogeny, \textcite{uriarte2004spatially} found that grouping based light guilds (akin to functional groups; \cite{blondel2003guilds}) produced a parsimonious model more frequently than grouping taxonimically (with both groupings defined \emph{a priori}).
Perhaps an attempt to produce functional groups based upon competitive interactions is warranted, applying the framework I present here to a set of species spanning multiple functional groups.

Alternatively, it should be noted that the success of \textcite{allesina2009food} in isolating groups using AIC occurred in much larger study systems, with 25 or more species.
Alas, the algorithmic complexity of the brute force method I employed in this study is abyssmal, preventing the application of the method to studies with more than seven or eight species.
As such, a better method would be required to test for grouping of species interactions among a larger set of species.

@@

All in all, the method of \emph{a posteriori} parsimony-based selection of groups has not proven to be greatly effective, at least on such a closely reated set of species.
However, the results suggest a number of pieces of advice which might be borne in mind whenever selecting groupings \emph{a priori} for use in species interaction models.
Firstly, there may be no single best grouping, but perhaps multiple rather distinct groupings which each perform similarly.
Secondly, phylogeny does not appear to provide appropriate groupings.
Lastly, one should not implicitly assume that species are grouped in the same way with regard to their effects upon other species and their response to them; distinct groupings of the two might well be considered.

>>>>

However, making any use of a model requires experimentally or observationally determining the values of each parameter in the model.
This poses a problem for species interaction models, due to the sheer number of parameters that must be determined.
Even the simplest models, with no environmental dependence, require one parameter for the interaction of each pair of species \parencite{novak2016characterizing}.
Hence the number of parameters scales quadratically with the number of species; over \num{100} parameters are required for a community of just \num{10} species.
Determining these parameters experimentally requires at least one experiment per parameter, and preferably more \parencite{hart2018quantify}.
This is expensive, and requires a lot of effort.
As such, these models are typically restricted to a small number of focal species \parencite{ovaskainen2017species}.
Real communities can contain hundreds of species \parencite{cornell1999unsaturation}, or even thousands for microbial communities assessed by genetic metabarcoding \parencite{warton2015variables}.
We need a way to reduce the number of parameters the models require, to regularise them.

@@

I propose that groups might be selected without regard to any \emph{a priori} definition, based purely upon which gives the most parsimonious model.
In this study, I investigate the usefulness and correctness of such a selection, using experimental \emph{Drosophila} fecundity data provided by \textcite{terry2021natural}.
I fit the Lotka--Volterra model, with grouped species, to find the grouping that gives the most parsimonious model.
I group species independently with regard to their effects upon other species and their responses to other species, to find whether the appropriate groupings might differ for these two cases.
Finally, I compare the resultant groupings with species phylogeny, to provide further investigation of the suitability of \emph{a priori} phylogenetic grouping.

@@

Species in the same column grouping were constrained to have the same competitive effect upon all other species, while species in the same column grouping were constrained to have the same competitive response to all other species.
Formally, if species $a$ and $b$ were in the same column grouping, $\alpha_{ia}=\alpha_{ib} \, \forall i$, and if $a$ and $b$ were in the same row grouping, $\alpha_{aj}=\alpha_{bj} \, \forall j$.
As such, the total number of free competition coefficients in the model was equal to the number of row groups times the number of column groups.
The species' intrinsic growth rates, $r_i$, were not affected by grouping; each species always had its own growth rate fitted independently of other species.

@@

To assess the parsimony of each model, all \num{\Sexpr{numbers::bell(6)^2}} possible models were fitted.
As a measure of parsimony, I selected AIC over BIC on the basis that we want to find the best model for making ecological predictions, rather than seeking the ``true'' \parencite{aho2014model}.
Further, AIC will be more conservative in promoting grouping, as BIC penalises number of parameters more for any reasonably sized data set.
As such, I calculated the AIC for each model.

@@

\begin{figure}
	\centering
	\begin{forest} phylogeny
		[
			[
				[
					[\emph{D. simulans}]
					[\emph{D. birchii}]
				]
				[
					[\emph{D. pandora}]
					[\emph{D. pseudoananassae}]
				]
			]
			[
				[
					[\emph{D. pallidifrons}]
					[\emph{D. sulfurigaster}]
				]
			]
		]
	\end{forest}
	%TODO: Mark subgroups and subgenera on the diagram.

@@

A high likelihood of coclassification was associated with phylogeny in one instance, but by no means consistently.
The phylogeny divides the six species into three most closely related pairs (\cref{fig:phylogeny}).
The pair of \emph{D. pallidifrons} and \emph{D. sulfurigaster}, which both belonged to a different subgenus than the other four species, were also the only pair of species to be coclassified with relatively high probability in both the row and column groupings.
In contrast, each of the other pairs were coclassified with very low probabilities in both the row and column groupings.
Furthermore, both row and column groupings gave high probabilities of coclassification to pairs of species split across the two subgenera in the study---the row groupings frequently coclassified \emph{D. sulfurigaster} and \emph{D. simulans}, while the column groupings frequently coclassified both \emph{D. pallidifrons} and \emph{D. sulfurigaster} of the Drosophila subgenus with \emph{D. birchii} of the Sophophora subgenus.

@@

This stands in contrast to the popular wisdom in ecology that species should form some sort of clear groups, be they the functional groups so often used \parencite{blondel2003guilds}, or taxonomic groups per a relationship between phylogeny and ecological niche \parencite{godoy2014phylogenetic}.
It also contrasts with previous studies that have succeeded in finding clear groups, such as \textcite{allesina2009food}, who used AIC to find the best grouping of species, but in a binary food web rather than a qualitative competitive context.
Perhaps species simply do not group to the same extent in a competitive context.
The results suggest that this is unlikely, as grouping was still likely to increase parsimony.
Further, there is no reason to suspect that species would not have clear similarities or dissimilarities in their ecological niches, and therefore their competitive effects and responses.
Perhaps the lack of clarity in the groupings is instead a product of the study system chosen.

I chose a study system of \emph{Drosophila} partially on the basis that it would allow clear analysis of the species phylogenetic relationships.
Phylogeny proved a poor predictor of optimal grouping, as previous studies have suggested that it might \parencite{godoy2014phylogenetic,cahill2008does}.
A study system with species spanning multiple functional groups, however, might have shown clearer groupings.
While functional groups have previously shown to be poor predictors of competitive interactions \parencite{bret2008plant}, studies in other contexts have succeeded in isolating functional groups by selecting for model parsimony \parencite{allesina2009food}.
In further support of functional groups over phylogeny, \textcite{uriarte2004spatially} found that grouping based light guilds (akin to functional groups; \cite{blondel2003guilds}) produced a parsimonious model more frequently than grouping taxonomically (with both groupings defined \emph{a priori}).
Perhaps an attempt to produce functional groups based upon competitive interactions is warranted, applying the framework I present here to a set of species spanning multiple functional groups.

Alternatively, it should be noted that the success of \textcite{allesina2009food} in isolating groups using AIC occurred in much larger study systems, with 25 or more species.
Alas, the algorithmic complexity of the brute force method I employed in this study is abysmal, preventing the application of the method to studies with more than seven or eight species.
As such, a better method would be required to test for grouping of species interactions among a larger set of species.

@@

All in all, the method of \emph{a posteriori} parsimony-based selection of groups has not proven to be greatly effective, at least on such a closely related set of species.
However, the results suggest a number of pieces of advice which might be borne in mind whenever selecting groupings \emph{a priori} for use in species interaction models.
Firstly, there may be no single best grouping, but perhaps multiple rather distinct groupings which each perform similarly.
Secondly, phylogeny does not appear to provide appropriate groupings.
Lastly, one should not implicitly assume that species are grouped in the same way with regard to their effects upon other species and their response to them; distinct groupings of the two might well be considered.
